---
title: "Analysis of Severe Weather Events on Public Health and Economy in Communities and Municipalities"
output:
  html_document:
    keep_md: false
    theme: united
---

# Severe Weather Effects on Public Health/Economy in Communities and Municipalities

## Synopsis

This is a synopsis in less than 10 complete sentences describing and
summarizing my analysis.

## Data Processing

### Dependency Loading

The libraries that are used throughout this document are installed if
necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

### Obtaining and loading source data

```{r cache=T}
temp <- tempfile()
data_url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(data_url,temp)
dat <- read.table(unz(temp, "stormdata.csv"),
                  na.strings = "NA", header=T, quote="\"", sep=",")
unlink(temp)
```

Describing in words and code how the data were loaded into R and processed for analysis.

- Start with raw CSV
- Cannot do any processing outside of document
- Justification and description for any transformations

## Results

- Must contain at least 1 figure
- Document can have no more than 3 figures
- Figures must have descriptive captions

### Events most harmful to population health

### Events with greatest economic consequences
