---
title: "Analysis of Severe Weather Events on Public Health and Economy in Communities and Municipalities"
output:
  html_document:
    keep_md: false
    theme: united
---

# Severe Weather Effects on Health/Economy in Communities and Municipalities

## Synopsis

The following is an analysis of the NOAA Storm Events data. The data begins in
the year 1950 and continues to 2011, with 902297 records. The data is composed
by 37 columns. Data will be cleansed, re-orderd and analyised in order to
answer the questions:

- Across the United States, which types of events are most harmful with respect
to population health?
- Across the United States, which types of events have the greatest economic
consequences?ummarizing my analysis.

In the end, we will show that Tornados cause an overwhelming amount of impact
on population health compared to other storm event types, causing 5633
fatalities and 91346 injuries.

Likewise, floods are the most impactful to economic health, causing around 150
billion USD in damage.

## Data Processing

### Dependency Loading

The libraries that are used throughout this document are installed if
necessary and then loaded by use of the following dependency function.

```{r libs, results="hide", warning=FALSE, message=FALSE, cache=TRUE}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

### Obtaining and loading source data

- [Storm Data][data] used in this analysis

To load this data, we request the raw CSV file from the remote location and
cache it locally. The data is then inspected at a high level by viewing its
head, and column names. This information was compared against the
documentation found in the following two sources.

- National Weather Service [Storm Data Documentation][docs]
- National Climatic Data Center Storm Events [FAQ][faq]

A high-level view of the data is included in the output below.

```{r read_data, cache=TRUE}
t <- "./storm.csv.bz2"
if (!file.exists(t)) {
    u <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormcsv.bz2'
    download.file(u,destfile=t, method="curl")
}
dat <- read.table(bzfile(t), na.strings = "NA", header=T, quote="\"", sep=",")
```

```{r summary, cache=TRUE}
colnames(dat);
head(dat);
```

#### Data transformations

For the following transformations, I'm loading and using the [dplyr][dplyr] and
[lubridate][lubridate] libraries.

```{r trans_depend, results="hide", warning=FALSE, message=FALSE}
depend('dplyr')
depend('lubridate')
```

Grouping all the data by event type then allows for easy summarization by
fatalities and injuries. I have sorted the data to show the most impactful
event types first. This data frame will be used to plot a figure analyzing the
events with the greatest effect on population health, below.

_Note: There are many variations of factors in the EVTYPE column. Some of these
variations appear to be referring to the same type of storm event, however,
they are not documented as such in the supporting documentation and have been
left as-is intentionally as a result._

```{r trans_pop_health}
pop_health <- dat %>%
    group_by(EVTYPE) %>%
    summarize(TOTAL_FATALITIES=sum(FATALITIES),
              TOTAL_INJURIES=sum(INJURIES)) %>%
    arrange(desc(TOTAL_FATALITIES), desc(TOTAL_INJURIES))

head(pop_health)
```

Calculating economic impact is a degree more difficult. There are several
fields related to this measurement: PROPDMG, PROPDMGEXP, CROPDMG, and
CROPDMGEXP. The PROPDMG and CROPDMG columns represent property and crop damages
in dollar values, respectively. The EXP columns give the order of magnitude of
those damages, ranging from thousands to billions. Unfortunately, these
exponent columns are very un-tidy, as demonstrated by listing their factor
levels:


```{r exp_factor}
levels(factor(dat$PROPDMGEXP))
```

To correct this, we need to carefully adjust all of these values to a single
numeric multiplier. Empty strings will get a 0 value. Symbols get a 1 value. H
(hundreds), K (thousands), M (millions), and B (billions) will get their
respective 2,3,6, or 9. Finally, to make sure our multiplication works, we'll
set any empty damage numbers to 0.

```{r tidy_trans_damage}
# replace missing data with 0's
dat$PROPDMGEXP <- as.character(dat$PROPDMGEXP)
dat$CROPDMGEXP <- as.character(dat$CROPDMGEXP)

dat$PROPDMGEXP[(dat$PROPDMGEXP == "")] <- 0
dat$PROPDMGEXP[(dat$PROPDMGEXP == "+") | (dat$PROPDMGEXP == "-") | (dat$PROPDMGEXP == "?")] <- 1
dat$PROPDMGEXP[(dat$PROPDMGEXP == "h") | (dat$PROPDMGEXP == "H")] <- 2
dat$PROPDMGEXP[(dat$PROPDMGEXP == "k") | (dat$PROPDMGEXP == "K")] <- 3
dat$PROPDMGEXP[(dat$PROPDMGEXP == "m") | (dat$PROPDMGEXP == "M")] <- 6
dat$PROPDMGEXP[(dat$PROPDMGEXP == "B")] <- 9

dat$CROPDMGEXP[(dat$CROPDMGEXP == "")] <- 0
dat$CROPDMGEXP[(dat$CROPDMGEXP == "+") | (dat$CROPDMGEXP == "-") | (dat$CROPDMGEXP == "?")] <- 1
dat$CROPDMGEXP[(dat$CROPDMGEXP == "h") | (dat$CROPDMGEXP == "H")] <- 2
dat$CROPDMGEXP[(dat$CROPDMGEXP == "k") | (dat$CROPDMGEXP == "K")] <- 3
dat$CROPDMGEXP[(dat$CROPDMGEXP == "m") | (dat$CROPDMGEXP == "M")] <- 6
dat$CROPDMGEXP[(dat$CROPDMGEXP == "B")] <- 9

dat$PROPDMGEXP <- as.integer(dat$PROPDMGEXP)
dat$CROPDMGEXP <- as.integer(dat$CROPDMGEXP)

dat$PROPDMG[(dat$PROPDMG == "")] <- 0
dat$CROPDMG[(dat$CROPDMG == "")] <- 0
```

```{r trans_damage}
trans_damage <- dat %>%
    mutate(TOTAL_DMG=(PROPDMG * 10^PROPDMGEXP + CROPDMG * 10^CROPDMGEXP)) %>%
    group_by(EVTYPE) %>%
    summarize(EVT_TOTAL_DMG=sum(TOTAL_DMG)) %>%
    arrange(desc(EVT_TOTAL_DMG))

head(trans_damage)
```

## Results

```{r graph_depend, results="hide", warning=FALSE, message=FALSE}
depend('ggplot2')
depend('ggthemes')
depend('scales')
```

### Events most harmful to population health

_Across the United States, which types of events (as indicated in the EVTYPE
variable) are most harmful with respect to population health?_

Plotting the top 10 event types by total fatalities and total injuries
clearly shows tornados as the leading impact to population health.

```{r pop_health_graph}
fatal_pop_health = head(pop_health %>% arrange(desc(TOTAL_FATALITIES)), 10)
ggplot(fatal_pop_health, aes(EVTYPE, TOTAL_FATALITIES) ) +
    geom_bar(stat = "identity") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Event Type", y="Total Fatalities", title="Total Fatalities by Storm Event Type") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5))

inj_pop_health = head(pop_health %>% arrange(desc(TOTAL_INJURIES)), 10)
ggplot(inj_pop_health, aes(EVTYPE, TOTAL_INJURIES) ) +
    geom_bar(stat = "identity") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Event Type", y="Total Injuries", title="Total Injuries by Storm Event Type") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5))

pop_health[pop_health$EVTYPE=="TORNADO",]
```

### Events with greatest economic consequences

_Across the United States, which types of events have the greatest economic consequences?_

When the damage numbers have been properly adjusted and combined, the top ten
most damaging event types

```{r trans_damage_graph}
trans_damage = head(trans_damage %>% arrange(desc(EVT_TOTAL_DMG)), 10)
ggplot(trans_damage, aes(EVTYPE, EVT_TOTAL_DMG) ) +
    geom_bar(stat = "identity") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Event Type", y="Total Damage in $", title="Total Economic Impact by Storm Event Type") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5))

trans_damage[trans_damage$EVTYPE=="FLOOD",]
```

- - - - -

[data]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormcsv.bz2 "Source Data"
[docs]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "Documentation"
[faq]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf "FAQ"
[dplyr]: https://cran.r-project.org/web/packages/dplyr/index.html "dplyr package"
[lubridate]: https://cran.r-project.org/web/packages/lubridate/index.html "lubridate package"
